---
title: "WGBS Quality Control: Dorsolateral Prefrontal Cortex (DLPFC)"
author: "HEART-GeN Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

## Learning Objectives

By the end of this analysis, you will be able to:

- Load and explore pre-generated WGBS data for the DLPFC
- Calculate and interpret key QC metrics for methylation data
- Identify samples and CpG sites that may need filtering
- Create publication-quality QC figures
- Make informed decisions about data quality thresholds

**Note:** For detailed background on WGBS methodology and QC concepts, see the companion tutorial `wgbs_qc_bsseq.qmd`.

---

## 1. Background

The **dorsolateral prefrontal cortex (DLPFC)** is a key brain region involved in executive functions, working memory, and cognitive control. DLPFC dysfunction has been consistently implicated in schizophrenia, making it a critical region for epigenetic studies.

In this analysis, we will perform quality control on WGBS data from DLPFC samples, focusing on:

- **Coverage**: Are CpG sites adequately sequenced?
- **Methylation distributions**: Do patterns look biologically plausible?
- **Sample correlations**: Are replicates similar to each other?
- **Principal components**: Are there technical batch effects?

---

## 2. Environment Setup

### Activate Conda Environment

Before running this analysis, ensure you have activated the lab's epigenomics environment:
```bash
source /projects/p32505/opt/miniforge3/etc/profile.d/conda.sh
conda activate /projects/p32505/opt/envs/epigenomics
```

### Load Libraries

```{r}
#| label: setup
#| message: false
library(here)
library(bsseq)
library(HDF5Array)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)
library(RColorBrewer)

## Set ggplot theme for consistency
theme_set(theme_bw(base_size = 12))

## Define region for this analysis
region <- "dlpfc"
region_name <- "DLPFC"
```

---

## 3. Data Loading

### Load Pre-Generated BSseq Object

The combined BSseq object was generated using the data loading scripts. This object contains methylation data for all chromosomes combined into a single, HDF5-backed structure for efficient access.

```{r}
#| label: load-data

output_dir <- here("qc_analysis/replication_cohort/_m")
if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
}

## Path to the combined data
data_path <- here("inputs/wgbs-data/combined/_m") # Explore "inputs/"

## Load the HDF5-backed BSseq object
bsseq_data <- loadHDF5SummarizedExperiment(
    file.path(data_path, paste0(region, "_bsseq_h5se"))
)

## Load phenotype data
phenotype_data <- read.csv(
    file.path(data_path, paste0(region, "_phenotypes.csv")),
    stringsAsFactors = FALSE
)

## Display basic information
cat("BSseq object summary:\n")
cat("  Number of CpG sites:", nrow(bsseq_data), "\n")
cat("  Number of samples:", ncol(bsseq_data), "\n")
cat("  Genome:", unique(genome(bsseq_data)), "\n")
```

### Phenotype Overview

We want to examine adults from our region of interest.

```{r}
#| label: phenotype-overview

## Display phenotype summary
phenotype_summary <- phenotype_data %>% 
                    filter(agedeath >= 17, region == !!region) %>%
                    group_by(primarydx) %>% 
                    summarize(
                        n = n(),
                        mean_age = round(mean(agedeath, na.rm = TRUE), 1),
                        sd_age = round(sd(agedeath, na.rm = TRUE), 1), 
                        n_female = sum(sex == "F", na.rm = TRUE),
                        n_male = sum(sex == "M", na.rm = TRUE),
                        n_AA = sum(race == "AA", na.rm = TRUE),
                        n_EA = sum(race == "EA", na.rm = TRUE),
                        .groups = "drop"
                    )
knitr::kable(phenotype_summary,
             col.names = c("Diagnosis", "N", "Mean Age", "SD Age", "Female", "Male", "African American", "European American"),
             caption = paste0("Sample demographics for ", region_name))
```

---

## 4. Sample Overview

### Sample Metadata Table

```{r}
#| label: sample-table

## Create sample summary table
sample_info <- phenotype_data %>%
    select(brnum, primarydx, agedeath, sex) %>%
    rename(
        Sample = brnum,
        Diagnosis = primarydx,
        Age = agedeath,
        Sex = sex
    )

## Display first few samples
knitr::kable(head(sample_info, 10),
             caption = "Sample metadata (first 10 samples)")
```

---

## 5. QC Analysis

### 5.1 Sequencing Coverage

Coverage refers to the number of sequencing reads that overlap each CpG site. Higher coverage provides more reliable methylation estimates.

#### Calculate Per-Sample Coverage

```{r}
#| label: coverage-calculation

## Get coverage matrix (reads per CpG per sample)
cov_matrix <- getCoverage(bsseq_data)
## Calculate mean coverage per sample
mean_coverage <- colMeans(cov_matrix, na.rm = TRUE)
## Create coverage summary dataframe
coverage_df <- tibble(
    Sample = names(mean_coverage),
    Mean_Coverage = mean_coverage
) %>%
    left_join(phenotype_data %>% select(brnum, primarydx),
              by = c("Sample" = "brnum"))
## Summary statistics (e.g., minimum, median, maxium)
cat("Coverage Summary:\n")
cat("  Minimum:", round(min(coverage_df$Mean_Coverage), 2), "x\n")
cat("  Median:", round(median(coverage_df$Mean_Coverage), 2), "x\n")
cat("  Maximum:", round(max(coverage_df$Mean_Coverage), 2), "x\n")
```

#### Coverage Barplot

```{r}
#| label: fig-coverage-barplot
#| fig-cap: "Mean sequencing coverage per sample. Samples below the red dashed line (5x coverage) may have unreliable methylation estimates."
#| fig-width: 10
#| fig-height: 6
ggsave(file = paste0(output_dir, "/plots/dlpfc_coverage_plot.pdf"), create.dir = TRUE, plot = ggplot(coverage_df, aes(x = reorder(Sample, Mean_Coverage), y = Mean_Coverage, fill = primarydx)) +
    geom_col() +
    geom_hline(yintercept = 5, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = paste0("Sequencing Coverage: ", region_name),
        x = "Sample",
        y = "Mean Coverage (reads per CpG)",
        fill = "Diagnosis"
    ) +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
        legend.position = "top"
    ) +
    scale_fill_brewer(palette = "Set2"))
```

#### Identify Low-Coverage Samples

```{r}
#| label: low-coverage-samples

## Flag samples with coverage < 5x
low_coverage_threshold <- 5
low_coverage_samples <- coverage_df %>%
    filter(Mean_Coverage < low_coverage_threshold)

if (nrow(low_coverage_samples) > 0) {
    cat("WARNING: The following samples have low coverage (<", low_coverage_threshold, "x):\n")
    print(low_coverage_samples)
} else {
    cat("All samples have adequate coverage (>=", low_coverage_threshold, "x)\n")
}
```

### 5.2 CpG Site Coverage

Not all CpG sites are equally well-covered. Sites with very low coverage across samples should be filtered before analysis.

```{r}
#| label: cpg-coverage

## Calculate coverage across CpGs
cpg_mean_cov <- rowMeans(cov_matrix, na.rm = TRUE)
## Calculate percentage of CpGs meeting thresholds
thresholds <- c(1, 3, 5, 10, 20)
cpg_coverage_summary <- sapply(thresholds, function(t) {
    mean(cpg_mean_cov >= t, na.rm = TRUE) * 100
})

coverage_table <- tibble(
    Threshold = paste0(">=", thresholds, "x"),
    Percent_CpGs = round(cpg_coverage_summary, 1)
)

## Output table using knitr::kable
knitr::kable(coverage_table,
             col.names = c("Coverage Threshold", "% CpGs Passing"),
             caption = "Percentage of CpG sites meeting coverage thresholds")
```

```{r}
#| label: fig-cpg-coverage-hist
#| fig-cap: "Distribution of mean coverage across CpG sites. Most sites should have adequate coverage (>5x)."
#| fig-width: 8
#| fig-height: 5

## Plot distribution (capped at 50x for visualization)
ggsave(file = paste0(output_dir, "/plots/dlpfc_cpg_coverage_plot.pdf"), plot = ggplot(tibble(Coverage = pmin(cpg_mean_cov, 50)), aes(x = Coverage)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
    geom_vline(xintercept = 5, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = paste0("CpG Coverage Distribution: ", region_name),
        x = "Mean Coverage (capped at 50x)",
        y = "Number of CpG Sites"
    ) +
    annotate("text", x = 5, y = Inf, label = "5x threshold",
             hjust = -0.1, vjust = 2, color = "red"))
```

### 5.3 Methylation Distribution

WGBS data typically shows a **bimodal distribution** with peaks near 0% (unmethylated) and 100% (fully methylated) methylation.

```{r}
#| label: methylation-extraction

## Extract methylation proportions (beta values)
## Use a subset of CpGs for efficiency (should be reproducible)
set.seed(123)
n_cpgs_sample <- min(100000, nrow(bsseq_data))
cpg_subset <- sample(nrow(bsseq_data), n_cpgs_sample)

meth_matrix <- as.matrix(getMeth(bsseq_data[cpg_subset, ], type = "raw"))
```

#### Per-Sample Methylation Histograms

```{r}
#| label: fig-meth-histograms
#| fig-cap: "Methylation distribution for selected samples. A bimodal pattern (peaks at 0 and 1) is expected for high-quality WGBS data."
#| fig-width: 10
#| fig-height: 8

## Select a few representative samples
sample_subset <- sample(colnames(meth_matrix), min(6, ncol(meth_matrix)))
## Prepare data for plotting
meth_long <- meth_matrix[, sample_subset] %>%
    as.data.frame() %>%
    pivot_longer(everything(), names_to = "Sample", values_to = "Methylation") %>%
    filter(!is.na(Methylation))
## Plot the data
ggsave(file = paste0(output_dir, "/plots/dlpfc_methylation_plot.pdf"), plot = ggplot(meth_long, aes(x = Methylation)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
    facet_wrap(~Sample, ncol = 3) +
    labs(
        title = paste0("Methylation Distributions: ", region_name),
        x = "Methylation Proportion (Beta)",
        y = "Count"
    ))
```

#### Density Plot by Diagnosis

```{r}
#| label: fig-meth-density
#| fig-cap: "Methylation density by diagnosis group. Distributions should be similar across groups if there are no technical artifacts."
#| fig-width: 8
#| fig-height: 5

## Calculate global methylation per sample
global_meth <- colMeans(meth_matrix, na.rm = TRUE)
## Create density data with diagnosis
meth_all <- meth_matrix %>%
    as.data.frame() %>%
    pivot_longer(everything(), names_to = "Sample", values_to = "Methylation") %>%
    left_join(phenotype_data %>% select(brnum, primarydx),
              by = c("Sample" = "brnum")) %>%
    filter(!is.na(Methylation))
## Sample for plotting efficiency
meth_sampled <- meth_all %>%
    group_by(primarydx) %>%
    slice_sample(n = 50000, replace = TRUE)
ggsave(file = paste0(output_dir, "/plots/dlpfc_meth_by_diagnosis_plot.pdf"), plot = ggplot(meth_sampled, aes(x = Methylation, color = primarydx)) +
    geom_density(linewidth = 1) +
    labs(
        title = paste0("Methylation Distribution by Diagnosis: ", region_name),
        x = "Methylation Proportion (Beta)",
        y = "Density",
        color = "Diagnosis"
    ) +
    scale_color_brewer(palette = "Set2"))
```

### 5.4 Sample Correlation

Pairwise correlations between samples help identify outliers. Biological replicates and similar sample types should have high correlations (typically >0.9).

```{r}
#| label: correlation-matrix

## Compute correlation matrix (on subset for efficiency)
cor_matrix <- cor(meth_matrix, use = "pairwise.complete.obs")
## Summary statistics
cor_values <- cor_matrix[upper.tri(cor_matrix)]
cat("Correlation Summary:\n")
cat("   Minimum:", round(min(cor_values), 3), "\n")
cat("   Median:", round(median(cor_values), 3), "\n")
cat("   Maximum:", round(max(cor_values), 3), "\n")
```



#### Identify Potential Outliers

```{r}
#| label: outlier-detection

## Calculate mean correlation for each sample
mean_cor <- rowMeans(cor_matrix) - 1/(ncol(cor_matrix))
## Flag samples with unusually low correlation
cor_threshold <- mean(mean_cor) - 2 * sd(mean_cor)

outlier_samples <- tibble(
    Sample = names(mean_cor),
    Mean_Correlation = mean_cor
) %>%
    filter(Mean_Correlation < cor_threshold)

if (nrow(outlier_samples) > 0) {
    cat("Potential outliers (mean correlation > 2 SD below mean):\n")
    print(outlier_samples)
} else {
    cat("No correlation-based outliers detected.\n")
}
```

### 5.5 Principal Component Analysis

PCA reduces the high-dimensional methylation data to a few principal components, helping visualize sample relationships and identify batch effects.

## 6. CpG Filtering

Based on the QC results, we apply coverage-based filtering to retain only high-quality CpG sites.
```{r}
#| label: cpg-filtering

## Define filtering criteria
min_coverage <- 5        # Minimum mean coverage per CpG
min_sample_frac <- 0.8   # CpG must be covered in at least 80% of samples

## Calculate coverage per CpG across samples
cov_per_cpg <- getCoverage(bsseq_data)
samples_with_coverage <- rowSums(cov_per_cpg >= min_coverage, na.rm = TRUE)
fraction_covered <- samples_with_coverage / ncol(bsseq_data)
## Apply filter
keep_cpgs <- fraction_covered >= min_sample_frac
## Summarize results (before and after filtering; how many CpGs were retained, rate of retention)
cat("CpG Filtering Summary:\n")
cat("  Original CpGs:", format(nrow(bsseq_data), big.mark = ","), "\n")
cat("  CpGs passing filter:", format(sum(keep_cpgs), big.mark = ","), "\n")
cat("  CpGs removed:", format(sum(!keep_cpgs), big.mark = ","), "\n")
cat("  Retention rate:", round(mean(keep_cpgs) * 100, 1), "%\n")

filtered_bsseq_data <- bsseq_data[keep_cpgs]
```

```{r}
#| label: pca-analysis
## Remove rows with missing values for PCA
# meth_matrix <- meth_matrix[keep_cpgs]
filtered_bsseq_data <- bsseq_data[keep_cpgs]
set.seed(123)
n_cpgs_sample <- min(100000, nrow(filtered_bsseq_data))
cpg_subset <- sample(nrow(filtered_bsseq_data), n_cpgs_sample)
meth_matrix <- as.matrix(getMeth(filtered_bsseq_data[cpg_subset, ], type = "raw"))


meth_complete <- meth_matrix[complete.cases(meth_matrix), ]
row_means_meth = rowMeans(meth_complete)
meth_complete <- meth_complete[row_means_meth > 0, ]
meth_complete <- unique(meth_complete)
# meth_complete <- Filter(function(row) length(unique(row)) != 1, meth_complete)
## Run PCA
pca_result <- prcomp(t(meth_complete), scale. = TRUE, center = TRUE)
## Calculate variance explained
var_explained <- summary(pca_result)$importance[2, ] * 100
## Print the top 3-5 PCs
cat("Variance explained by PCs:\n")
cat("  PC1:", round(var_explained[1], 1), "%\n")
cat("  PC2:", round(var_explained[2], 1), "%\n")
cat("  PC3:", round(var_explained[3], 1), "%\n")
```

```{r}
#| label: fig-pca-diagnosis
#| fig-cap: "PCA colored by diagnosis. Separation along PC1/PC2 may indicate biological differences or technical artifacts."
#| fig-width: 8
#| fig-height: 6

## Prepare PCA data for plotting
pca_df <- as.data.frame(pca_result$x) %>%
    rownames_to_column("Sample") %>%
    left_join(phenotype_data %>% select(brnum, primarydx, sex, agedeath),
              by = c("Sample" = "brnum"))

```




## 7. QC Summary Table

```{r}
#| label: qc-summary-table

## Compile QC metrics for each sample
qc_summary <- tibble(
    Sample = colnames(bsseq_data),
    Mean_Coverage = mean_coverage[colnames(bsseq_data)],
    Mean_Correlation = mean_cor[colnames(bsseq_data)],
    Global_Methylation = global_meth[colnames(bsseq_data)]
) %>%
    left_join(phenotype_data %>% select(brnum, primarydx, sex, agedeath),
              by = c("Sample" = "brnum")) %>%
    mutate(
        Coverage_Flag = ifelse(Mean_Coverage < 5, "Low", "OK"),
        Correlation_Flag = ifelse(Mean_Correlation < cor_threshold, "Outlier", "OK"),
        QC_Status = case_when(
            Coverage_Flag == "Low" | Correlation_Flag == "Outlier" ~ "Review",
            TRUE ~ "Pass"
        )
    )

## Display summary
knitr::kable(
    qc_summary %>%
        select(Sample, primarydx, Mean_Coverage, Mean_Correlation, QC_Status) %>%
        mutate(
            Mean_Coverage = round(Mean_Coverage, 1),
            Mean_Correlation = round(Mean_Correlation, 3)
        ),
    caption = "QC Summary Table"
)
```

### Samples Requiring Review

```{r}
#| label: samples-review

## What are the samples that require review
review_samples <- qc_summary %>% filter(QC_Status == "Review")

if (nrow(review_samples) > 0) {
    cat("The following samples may require additional review:\n\n")
    print(review_samples %>% select(Sample, primarydx, Mean_Coverage, Mean_Correlation, Coverage_Flag, Correlation_Flag))
} else {
    cat("All samples passed QC. No samples require additional review.\n")
}
```

---

## 8. Save QC Results

```{r}
#| label: save-results

## Create output directory


## Save QC summary table
write.csv(qc_summary,
          file.path(output_dir, paste0("qc_summary_", region, ".csv")),
          row.names = FALSE)
## Save filtered BSseq object
bsseq_filtered <- bsseq_data[keep_cpgs, ]
out_h5se <- file.path(output_dir, paste0(region, "_bsseq_qc_filtered_h5se"))
if (dir.exists(out_h5se)) unlink(out_h5se, recursive = TRUE)
saveHDF5SummarizedExperiment(bsseq_filtered, dir = out_h5se, replace = TRUE)

cat("Results saved to:", output_dir, "\n")
cat("  - QC summary: qc_summary_", region, ".csv\n", sep = "")
cat("  - Filtered BSseq: ", region, "_bsseq_qc_filtered_h5se/\n", sep = "")
```



```{r}
# Test all PCs, spearman correlation
# rho, p-value
# heatmap with all principal components, x is covariants, y is PCs
# cor.test

# make a csv to store the principal component dataframe, phenotype_data, so I don't have to rerun everytime

# relevel PCs


covariant_vec <- c("agedeath", "pmi", "snpPC1", "snpPC2", "snpPC3", "ph", "age_onset_schizo")
pc_vec = c()
rho_vec = c()

fdr_vec = c()
# list
for (pc in 1:20) {
  p_vec = c()
  for (covariant in covariant_vec) {
    pc_name <- paste0("PC", pc)
    pc_vec = c(pc_vec, pc_name)
    
    result <- cor.test(phenotype_data[[covariant]], pca_df[[pc_name]], method = "spearm", exact = FALSE)
    p_value <- result$p.value
    rho <- result$estimate[["rho"]]
    
    p_vec = c(p_vec, p_value)
    rho_vec = c(rho_vec, rho)
    
    
  }
  fdr_vec <- c(fdr_vec, p.adjust(p_vec, method = "fdr"))
  
}

df = data.frame(
     PC = pc_vec,
     covariant = covariant_vec,
     rho = rho_vec,
     p = p_vec,
     fdr = fdr_vec,
     stringsAsFactors = FALSE
)

# Trend level significance

# Make submission script step1.sh
# sbatch in _m
# in git commit message (closes #2)

pp = ggplot(df, aes(y = reorder(PC, as.numeric(sub("PC", "", PC))), x = covariant, label = ifelse(fdr < 0.05, "*", ""), fill = rho)) + geom_tile(color = "gray") + scale_fill_gradient2(low = "blue", mid = "white", high = "red", name = "rho value") + geom_text(color = "black", size = 7) + labs(y = "PC", x = "Technical and Biological Covariates")

ggsave(file = paste0(output_dir, "/plots/dlpfc_all_covariates_correlation_heapmap.pdf"), plot = pp, width = 10)

pp = ggplot(df, aes(y = reorder(PC, as.numeric(sub("PC", "", PC))), x = covariant, label = ifelse(fdr < 0.10, "*", ""), fill = rho)) + geom_tile(color = "gray") + scale_fill_gradient2(low = "blue", mid = "white", high = "red", name = "rho value") + geom_text(color = "black", size = 7) + labs(y = "PC", x = "Technical and Biological Covariates")

ggsave(file = paste0(output_dir, "/plots/dlpfc_all_covariates_trend_heapmap.pdf"), plot = pp, width = 10)


# wide and long figure
# organize into 1 plot
# flip plot

```

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
