---
title: "WGBS Quality Control: Multi-Region Comparison"
author: "HEART-GeN Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

## Learning Objectives

By the end of this analysis, you will be able to:

- Compare methylation patterns across multiple brain regions
- Identify samples present in multiple regions
- Assess data consistency for samples measured across regions
- Detect potential sample swaps or mislabeling

---

## 1. Background

This analysis compares WGBS data across three brain regions:

- **Caudate nucleus**: Subcortical structure involved in motor control and reward
- **DLPFC**: Prefrontal region involved in executive functions
- **Hippocampus**: Memory and learning center

Comparing data across regions helps us:

1. Understand region-specific methylation patterns
2. Verify sample identity for individuals with data from multiple regions
3. Identify potential technical issues that affect specific regions

---

## 2. Environment Setup

### Activate Conda Environment

```bash
source /projects/p32505/opt/miniforge3/etc/profile.d/conda.sh
conda activate /projects/p32505/opt/envs/epigenomics
```

### Load Libraries

```{r}
#| label: setup
#| message: false

library(here)
library(bsseq)
library(HDF5Array)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(ggvenn)
library(RColorBrewer)

## Set ggplot theme
theme_set(theme_bw(base_size = 12))

## Define regions
regions <- c("caudate", "dlpfc", "hippocampus")
region_names <- c(
    caudate = "Caudate Nucleus",
    dlpfc = "DLPFC",
    hippocampus = "Hippocampus"
)
```

---

## 3. Data Loading

### Load BSseq Objects for All Regions

```{r}
#| label: load-data

## Path to combined data
data_path <- here("inputs/wgbs-data/combined/_m")

## Load all BSseq objects
bsseq_list <- list()
pheno_list <- list()

for (reg in regions) {
    cat("Loading", reg, "...\n")

    ## Load BSseq object
    h5se_path <- file.path(data_path, paste0(reg, "_bsseq_h5se"))
    if (dir.exists(h5se_path)) {
        bsseq_list[[reg]] <- loadHDF5SummarizedExperiment(h5se_path)
        cat("  Samples:", ncol(bsseq_list[[reg]]), "\n")
        cat("  CpG sites:", nrow(bsseq_list[[reg]]), "\n")
    } else {
        cat("  WARNING: Data not found for", reg, "\n")
    }

    ## Load phenotype data
    pheno_path <- file.path(data_path, paste0(reg, "_phenotypes.csv"))
    if (file.exists(pheno_path)) {
        pheno_list[[reg]] <- read.csv(pheno_path, stringsAsFactors = FALSE)
    }
}
```

### Summary Table

```{r}
#| label: region-summary

## Create summary table
region_summary <- tibble(
    Region = region_names[regions],
    Samples = sapply(bsseq_list, ncol),
    CpG_Sites = sapply(bsseq_list, nrow)
)

knitr::kable(region_summary,
             col.names = c("Brain Region", "N Samples", "N CpG Sites"),
             caption = "Summary of WGBS data by brain region",
             format.args = list(big.mark = ","))
```

---

## 4. Sample Overlap Analysis

### 4.1 Venn Diagram of Samples

Many donors may have samples from multiple brain regions. Understanding this overlap is crucial for paired analyses.

```{r}
#| label: fig-venn-samples
#| fig-cap: "Venn diagram showing sample overlap across brain regions. Samples are identified by donor ID (brnum)."
#| fig-width: 8
#| fig-height: 6

## Get sample lists for each region
sample_lists <- lapply(names(bsseq_list), function(reg) {
    colnames(bsseq_list[[reg]])
})
names(sample_lists) <- region_names[names(bsseq_list)]

## Create Venn diagram
ggvenn(sample_lists,
       fill_color = brewer.pal(3, "Set2"),
       stroke_size = 0.5,
       set_name_size = 4)
```

### 4.2 Sample Overlap Table

```{r}
#| label: sample-overlap-table

## Identify samples in multiple regions
all_samples <- unique(unlist(sample_lists))

## Count regions per sample
sample_region_counts <- sapply(all_samples, function(s) {
    sum(sapply(sample_lists, function(sl) s %in% sl))
})

## Create detailed table
overlap_df <- tibble(
    Sample = all_samples,
    N_Regions = sample_region_counts,
    In_Caudate = all_samples %in% sample_lists[["Caudate Nucleus"]],
    In_DLPFC = all_samples %in% sample_lists[["DLPFC"]],
    In_Hippocampus = all_samples %in% sample_lists[["Hippocampus"]]
) %>%
    arrange(desc(N_Regions), Sample)

## Summary
cat("Sample distribution across regions:\n")
table(sample_region_counts) %>% print()
```

```{r}
#| label: multi-region-samples

## Show samples present in 2+ regions
multi_region_samples <- overlap_df %>%
    filter(N_Regions >= 2)

if (nrow(multi_region_samples) > 0) {
    cat("\nSamples present in multiple regions:\n")
    knitr::kable(head(multi_region_samples, 20),
                 caption = paste0("Samples in 2+ regions (showing first 20 of ",
                                  nrow(multi_region_samples), ")"))
} else {
    cat("\nNo samples present in multiple regions.\n")
}
```

---

## 5. Coverage Comparison

### 5.1 Coverage by Region

```{r}
#| label: coverage-by-region

## Calculate mean coverage for each sample in each region
coverage_data <- lapply(names(bsseq_list), function(reg) {
    cov <- getCoverage(bsseq_list[[reg]])
    mean_cov <- colMeans(cov, na.rm = TRUE)
    tibble(
        Sample = names(mean_cov),
        Mean_Coverage = mean_cov,
        Region = region_names[reg]
    )
}) %>%
    bind_rows()

## Summary statistics by region
coverage_summary <- coverage_data %>%
    group_by(Region) %>%
    summarise(
        N = n(),
        Mean = round(mean(Mean_Coverage), 1),
        SD = round(sd(Mean_Coverage), 1),
        Median = round(median(Mean_Coverage), 1),
        Min = round(min(Mean_Coverage), 1),
        Max = round(max(Mean_Coverage), 1),
        .groups = "drop"
    )

knitr::kable(coverage_summary,
             caption = "Coverage statistics by brain region")
```

```{r}
#| label: fig-coverage-boxplot
#| fig-cap: "Distribution of sequencing coverage across brain regions. The red dashed line indicates the 5x minimum coverage threshold."
#| fig-width: 8
#| fig-height: 5

ggplot(coverage_data, aes(x = Region, y = Mean_Coverage, fill = Region)) +
    geom_boxplot(alpha = 0.7, outlier.alpha = 0.5) +
    geom_jitter(width = 0.2, alpha = 0.4, size = 1) +
    geom_hline(yintercept = 5, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = "Sequencing Coverage by Brain Region",
        x = "Brain Region",
        y = "Mean Coverage (reads per CpG)"
    ) +
    scale_fill_brewer(palette = "Set2") +
    theme(legend.position = "none")
```

### 5.2 Coverage Consistency for Paired Samples

For samples with data in multiple regions, we can check if coverage is consistent.

```{r}
#| label: fig-paired-coverage
#| fig-cap: "Coverage comparison for samples with data in multiple regions. Points should cluster around the diagonal if coverage is consistent."
#| fig-width: 10
#| fig-height: 4

## Get samples in multiple regions
multi_samples <- overlap_df$Sample[overlap_df$N_Regions >= 2]

if (length(multi_samples) > 0) {
    ## Prepare paired coverage data
    paired_coverage <- coverage_data %>%
        filter(Sample %in% multi_samples) %>%
        pivot_wider(names_from = Region, values_from = Mean_Coverage)

    ## Create comparison plots
    plot_list <- list()

    region_pairs <- list(
        c("Caudate Nucleus", "DLPFC"),
        c("Caudate Nucleus", "Hippocampus"),
        c("DLPFC", "Hippocampus")
    )

    for (pair in region_pairs) {
        if (all(pair %in% names(paired_coverage))) {
            p <- paired_coverage %>%
                filter(!is.na(.data[[pair[1]]]) & !is.na(.data[[pair[2]]])) %>%
                ggplot(aes(x = .data[[pair[1]]], y = .data[[pair[2]]])) +
                geom_point(alpha = 0.6, size = 2) +
                geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "red") +
                labs(
                    title = paste(pair[1], "vs", pair[2]),
                    x = paste(pair[1], "Coverage"),
                    y = paste(pair[2], "Coverage")
                ) +
                coord_fixed()
            plot_list[[length(plot_list) + 1]] <- p
        }
    }

    if (length(plot_list) > 0) {
        cowplot::plot_grid(plotlist = plot_list, ncol = 3)
    }
} else {
    cat("No samples in multiple regions to compare.\n")
}
```

---

## 6. Global Methylation Comparison

### 6.1 Global Methylation by Region

```{r}
#| label: global-methylation

## Calculate global methylation for each sample (using subset for efficiency)
set.seed(123)

global_meth_data <- lapply(names(bsseq_list), function(reg) {
    bs <- bsseq_list[[reg]]

    ## Sample CpGs for efficiency
    n_sample <- min(50000, nrow(bs))
    idx <- sample(nrow(bs), n_sample)

    meth <- getMeth(bs[idx, ], type = "raw")
    global_meth <- colMeans(meth, na.rm = TRUE)

    tibble(
        Sample = names(global_meth),
        Global_Methylation = global_meth,
        Region = region_names[reg]
    )
}) %>%
    bind_rows()

## Summary
meth_summary <- global_meth_data %>%
    group_by(Region) %>%
    summarise(
        N = n(),
        Mean = round(mean(Global_Methylation, na.rm = TRUE), 3),
        SD = round(sd(Global_Methylation, na.rm = TRUE), 3),
        .groups = "drop"
    )

knitr::kable(meth_summary,
             caption = "Global methylation levels by brain region")
```

```{r}
#| label: fig-global-meth-density
#| fig-cap: "Distribution of global methylation levels by brain region. Systematic differences may indicate biological variation or technical effects."
#| fig-width: 8
#| fig-height: 5

ggplot(global_meth_data, aes(x = Global_Methylation, fill = Region, color = Region)) +
    geom_density(alpha = 0.4, linewidth = 1) +
    labs(
        title = "Global Methylation Distribution by Region",
        x = "Global Methylation (mean beta)",
        y = "Density"
    ) +
    scale_fill_brewer(palette = "Set2") +
    scale_color_brewer(palette = "Set2")
```

### 6.2 Paired Sample Methylation Consistency

```{r}
#| label: fig-paired-methylation
#| fig-cap: "Global methylation comparison for samples with data in multiple regions. Strong correlation indicates sample consistency."
#| fig-width: 10
#| fig-height: 4

if (length(multi_samples) > 0) {
    ## Prepare paired methylation data
    paired_meth <- global_meth_data %>%
        filter(Sample %in% multi_samples) %>%
        pivot_wider(names_from = Region, values_from = Global_Methylation)

    ## Create comparison plots
    plot_list <- list()

    for (pair in region_pairs) {
        if (all(pair %in% names(paired_meth))) {
            sub_data <- paired_meth %>%
                filter(!is.na(.data[[pair[1]]]) & !is.na(.data[[pair[2]]]))

            if (nrow(sub_data) >= 3) {
                cor_val <- cor(sub_data[[pair[1]]], sub_data[[pair[2]]],
                               use = "complete.obs")

                p <- ggplot(sub_data, aes(x = .data[[pair[1]]], y = .data[[pair[2]]])) +
                    geom_point(alpha = 0.6, size = 2) +
                    geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
                    labs(
                        title = paste(pair[1], "vs", pair[2]),
                        subtitle = paste0("r = ", round(cor_val, 3)),
                        x = paste(pair[1], "Methylation"),
                        y = paste(pair[2], "Methylation")
                    ) +
                    coord_fixed()
                plot_list[[length(plot_list) + 1]] <- p
            }
        }
    }

    if (length(plot_list) > 0) {
        cowplot::plot_grid(plotlist = plot_list, ncol = 3)
    }
} else {
    cat("No samples in multiple regions to compare.\n")
}
```

---

## 7. Sample Identity Check

For samples present in multiple regions from the same donor, their methylation profiles should be correlated. Low correlation may indicate sample swaps.

```{r}
#| label: sample-identity-check

if (length(multi_samples) >= 2) {
    ## Find common CpG sites across all regions
    ## This requires all BSseq objects to be in the same coordinate system

    cat("Checking sample identity consistency across regions...\n\n")

    ## Calculate within-donor correlation for paired samples
    identity_results <- list()

    for (sample_id in multi_samples) {
        ## Get regions this sample is in
        sample_regions <- names(bsseq_list)[sapply(bsseq_list, function(bs) {
            sample_id %in% colnames(bs)
        })]

        if (length(sample_regions) >= 2) {
            ## Store global methylation values for this sample
            sample_meth <- global_meth_data %>%
                filter(Sample == sample_id)

            identity_results[[sample_id]] <- sample_meth
        }
    }

    ## Identify potential issues (samples with very different methylation across regions)
    paired_meth_check <- global_meth_data %>%
        filter(Sample %in% multi_samples) %>%
        group_by(Sample) %>%
        summarise(
            N_Regions = n(),
            Meth_Range = max(Global_Methylation) - min(Global_Methylation),
            .groups = "drop"
        ) %>%
        arrange(desc(Meth_Range))

    ## Flag samples with large methylation differences
    threshold <- 0.05  # 5% difference
    flagged <- paired_meth_check %>%
        filter(Meth_Range > threshold)

    if (nrow(flagged) > 0) {
        cat("WARNING: The following samples show large methylation differences across regions:\n")
        cat("(This may indicate technical issues or true biological variation)\n\n")
        print(flagged)
    } else {
        cat("All paired samples show consistent global methylation across regions.\n")
    }
} else {
    cat("Insufficient paired samples for identity check.\n")
}
```

---

## 8. CpG Site Overlap

### Compare CpG sites covered across regions

```{r}
#| label: cpg-overlap

## Get CpG coordinates for each region
cpg_coords <- lapply(names(bsseq_list), function(reg) {
    gr <- rowRanges(bsseq_list[[reg]])
    paste0(seqnames(gr), ":", start(gr))
})
names(cpg_coords) <- region_names[names(bsseq_list)]

## Count overlaps
n_cpgs <- sapply(cpg_coords, length)
common_all <- Reduce(intersect, cpg_coords)

cat("CpG Site Summary:\n")
cat("  Caudate:", format(n_cpgs[1], big.mark = ","), "sites\n")
cat("  DLPFC:", format(n_cpgs[2], big.mark = ","), "sites\n")
cat("  Hippocampus:", format(n_cpgs[3], big.mark = ","), "sites\n")
cat("  Common to all:", format(length(common_all), big.mark = ","), "sites\n")
```

---

## 9. Summary

```{r}
#| label: summary-table

## Compile cross-region summary
summary_table <- tibble(
    Metric = c(
        "Total unique samples",
        "Samples in 1 region",
        "Samples in 2 regions",
        "Samples in all 3 regions",
        "Mean coverage (overall)",
        "Mean global methylation"
    ),
    Value = c(
        length(all_samples),
        sum(sample_region_counts == 1),
        sum(sample_region_counts == 2),
        sum(sample_region_counts == 3),
        paste0(round(mean(coverage_data$Mean_Coverage), 1), "x"),
        round(mean(global_meth_data$Global_Methylation, na.rm = TRUE), 3)
    )
)

knitr::kable(summary_table,
             caption = "Multi-Region Analysis Summary")
```

---

## 10. Save Results

```{r}
#| label: save-results

## Create output directory
output_dir <- here("qc_analysis/_m")
if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
}

## Save sample overlap information
write.csv(overlap_df,
          file.path(output_dir, "sample_overlap_multiregion.csv"),
          row.names = FALSE)

## Save coverage comparison
write.csv(coverage_data,
          file.path(output_dir, "coverage_by_region.csv"),
          row.names = FALSE)

## Save global methylation data
write.csv(global_meth_data,
          file.path(output_dir, "global_methylation_by_region.csv"),
          row.names = FALSE)

cat("Results saved to:", output_dir, "\n")
```

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
