---
title: "WGBS Quality Control: Caudate Nucleus"
author: "HEART-GeN Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

## Learning Objectives

By the end of this analysis, you will be able to:

- Load and explore pre-generated WGBS data for the caudate nucleus
- Calculate and interpret key QC metrics for methylation data
- Identify samples and CpG sites that may need filtering
- Create publication-quality QC figures
- Make informed decisions about data quality thresholds

**Note:** For detailed background on WGBS methodology and QC concepts, see the companion tutorial `wgbs_qc_bsseq.qmd`.

**Please review `wgbs_qc_bsseq.qmd` first for background on WGBS and QC concepts.**

---

## 1. Background

The **caudate nucleus** is a subcortical brain structure involved in motor control, learning, and reward processing. Alterations in DNA methylation in the caudate have been associated with neuropsychiatric disorders including schizophrenia.

In this analysis, we will perform quality control on WGBS data from caudate samples, focusing on:

- **Coverage**: Are CpG sites adequately sequenced?
- **Methylation distributions**: Do patterns look biologically plausible?
- **Sample correlations**: Are replicates similar to each other?
- **Principal components**: Are there technical batch effects?

---

## 2. Environment Setup

### Activate Conda Environment

Before running this analysis, ensure you have activated the lab's epigenomics environment:
```bash
source /projects/p32505/opt/miniforge3/etc/profile.d/conda.sh
conda activate /projects/p32505/opt/envs/epigenomics
```

### Load Libraries

```{r}
#| label: setup
#| message: false

library(here)
library(bsseq)
library(HDF5Array)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)
library(RColorBrewer)

## Set ggplot theme for consistency
theme_set(theme_bw(base_size = 12))

## Define region for this analysis
region <- "caudate"
region_name <- "Caudate Nucleus"
```

---

## 3. Data Loading

### Load Pre-Generated BSseq Object

The combined BSseq object was generated using the data loading scripts. This object contains methylation data for all chromosomes combined into a single, HDF5-backed structure for efficient access.

```{r}
#| label: load-data

## Path to the combined data
data_path <- here("inputs/wgbs-data/combined/_m")

## Load the HDF5-backed BSseq object
bsseq_data <- loadHDF5SummarizedExperiment(
    file.path(data_path, paste0(region, "_bsseq_h5se"))
)

## Load phenotype data
phenotype_data <- read.csv(
    file.path(data_path, paste0(region, "_phenotypes.csv")),
    stringsAsFactors = FALSE
)

## Display basic information
cat("BSseq object summary:\n")
cat("  Number of CpG sites:", nrow(bsseq_data), "\n")
cat("  Number of samples:", ncol(bsseq_data), "\n")
cat("  Genome:", unique(genome(bsseq_data)), "\n")
```

### Phenotype Overview

We want to examine adults from our region of interest.

```{r}
#| label: phenotype-overview

## Display phenotype summary
phenotype_summary <- phenotype_data %>%
    filter(agedeath >= 17, region == !!region) %>%
    group_by(dx) %>%
    summarise(
        n = n(),
        mean_age = round(mean(agedeath, na.rm = TRUE), 1),
        sd_age = round(sd(agedeath, na.rm = TRUE), 1),
        n_female = sum(sex == "F", na.rm = TRUE),
        n_male = sum(sex == "M", na.rm = TRUE),
        .groups = "drop"
    )

knitr::kable(phenotype_summary,
             col.names = c("Diagnosis", "N", "Mean Age", "SD Age", "Female", "Male"),
             caption = paste0("Sample demographics for ", region_name))
```

---

## 4. Sample Overview

### Sample Metadata Table

```{r}
#| label: sample-table

## Create sample summary table
sample_info <- phenotype_data %>%
    select(brnum, dx, agedeath, sex) %>%
    rename(
        Sample = brnum,
        Diagnosis = dx,
        Age = agedeath,
        Sex = sex
    )

## Display first few samples
knitr::kable(head(sample_info, 10),
             caption = "Sample metadata (first 10 samples)")
```

---

## 5. QC Analysis

### 5.1 Sequencing Coverage

Coverage refers to the number of sequencing reads that overlap each CpG site. Higher coverage provides more reliable methylation estimates.

#### Calculate Per-Sample Coverage

```{r}
#| label: coverage-calculation

## Get coverage matrix (reads per CpG per sample)
cov_matrix <- getCoverage(bsseq_data)

## Calculate mean coverage per sample
mean_coverage <- colMeans(cov_matrix, na.rm = TRUE)

## Create coverage summary dataframe
coverage_df <- tibble(
    Sample = names(mean_coverage),
    Mean_Coverage = mean_coverage
) %>%
    left_join(phenotype_data %>% select(brnum, dx),
              by = c("Sample" = "brnum"))

## Summary statistics
### This is just a couple of things you can look at with nice rounding.
cat("Coverage Summary:\n")
cat("  Minimum:", round(min(coverage_df$Mean_Coverage), 2), "x\n")
cat("  Median:", round(median(coverage_df$Mean_Coverage), 2), "x\n")
cat("  Maximum:", round(max(coverage_df$Mean_Coverage), 2), "x\n")
```

#### Coverage Barplot

```{r}
#| label: fig-coverage-barplot
#| fig-cap: "Mean sequencing coverage per sample. Samples below the red dashed line (5x coverage) may have unreliable methylation estimates."
#| fig-width: 10
#| fig-height: 6

ggplot(coverage_df, aes(x = reorder(Sample, Mean_Coverage), y = Mean_Coverage, fill = dx)) +
    geom_col() +
    geom_hline(yintercept = 5, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = paste0("Sequencing Coverage: ", region_name),
        x = "Sample",
        y = "Mean Coverage (reads per CpG)",
        fill = "Diagnosis"
    ) +
    theme(
        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, size = 6),
        legend.position = "top"
    ) +
    scale_fill_brewer(palette = "Set2")
```

#### Identify Low-Coverage Samples

```{r}
#| label: low-coverage-samples

## Flag samples with coverage < 5x
low_coverage_threshold <- 5
low_coverage_samples <- coverage_df %>%
    filter(Mean_Coverage < low_coverage_threshold)

if (nrow(low_coverage_samples) > 0) {
    cat("WARNING: The following samples have low coverage (<", low_coverage_threshold, "x):\n")
    print(low_coverage_samples)
} else {
    cat("All samples have adequate coverage (>=", low_coverage_threshold, "x)\n")
}
```

### 5.2 CpG Site Coverage

Not all CpG sites are equally well-covered. Sites with very low coverage across samples should be filtered before analysis.

```{r}
#| label: cpg-coverage

## Calculate coverage across CpGs
cpg_mean_cov <- rowMeans(cov_matrix, na.rm = TRUE)

## Calculate percentage of CpGs meeting thresholds
thresholds <- c(1, 3, 5, 10, 20)
cpg_coverage_summary <- sapply(thresholds, function(t) {
    mean(cpg_mean_cov >= t, na.rm = TRUE) * 100
})

coverage_table <- tibble(
    Threshold = paste0(">=", thresholds, "x"),
    Percent_CpGs = round(cpg_coverage_summary, 1)
)

knitr::kable(coverage_table,
             col.names = c("Coverage Threshold", "% CpGs Passing"),
             caption = "Percentage of CpG sites meeting coverage thresholds")
```

```{r}
#| label: fig-cpg-coverage-hist
#| fig-cap: "Distribution of mean coverage across CpG sites. Most sites should have adequate coverage (>5x)."
#| fig-width: 8
#| fig-height: 5

## Plot distribution (capped at 50x for visualization)
ggplot(tibble(Coverage = pmin(cpg_mean_cov, 50)), aes(x = Coverage)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
    geom_vline(xintercept = 5, linetype = "dashed", color = "red", linewidth = 1) +
    labs(
        title = paste0("CpG Coverage Distribution: ", region_name),
        x = "Mean Coverage (capped at 50x)",
        y = "Number of CpG Sites"
    ) +
    annotate("text", x = 5, y = Inf, label = "5x threshold",
             hjust = -0.1, vjust = 2, color = "red")
```

### 5.3 Methylation Distribution

WGBS data typically shows a **bimodal distribution** with peaks near 0% (unmethylated) and 100% (fully methylated) methylation.

```{r}
#| label: methylation-extraction

## Extract methylation proportions (beta values)
## Use a subset of CpGs for efficiency
set.seed(123)
n_cpgs_sample <- min(100000, nrow(bsseq_data))
cpg_subset <- sample(nrow(bsseq_data), n_cpgs_sample)

meth_matrix <- getMeth(bsseq_data[cpg_subset, ], type = "raw")
```

#### Per-Sample Methylation Histograms

```{r}
#| label: fig-meth-histograms
#| fig-cap: "Methylation distribution for selected samples. A bimodal pattern (peaks at 0 and 1) is expected for high-quality WGBS data."
#| fig-width: 10
#| fig-height: 8

## Select a few representative samples
sample_subset <- sample(colnames(meth_matrix), min(6, ncol(meth_matrix)))

## Prepare data for plotting
meth_long <- meth_matrix[, sample_subset] %>%
    as.data.frame() %>%
    pivot_longer(everything(), names_to = "Sample", values_to = "Methylation") %>%
    filter(!is.na(Methylation))

ggplot(meth_long, aes(x = Methylation)) +
    geom_histogram(bins = 50, fill = "steelblue", color = "black", alpha = 0.7) +
    facet_wrap(~Sample, ncol = 3) +
    labs(
        title = paste0("Methylation Distributions: ", region_name),
        x = "Methylation Proportion (Beta)",
        y = "Count"
    )
```

#### Density Plot by Diagnosis

```{r}
#| label: fig-meth-density
#| fig-cap: "Methylation density by diagnosis group. Distributions should be similar across groups if there are no technical artifacts."
#| fig-width: 8
#| fig-height: 5

## Calculate global methylation per sample
global_meth <- colMeans(meth_matrix, na.rm = TRUE)

## Create density data with diagnosis
meth_all <- meth_matrix %>%
    as.data.frame() %>%
    pivot_longer(everything(), names_to = "Sample", values_to = "Methylation") %>%
    left_join(phenotype_data %>% select(brnum, dx),
              by = c("Sample" = "brnum")) %>%
    filter(!is.na(Methylation))

## Sample for plotting efficiency
meth_sampled <- meth_all %>%
    group_by(dx) %>%
    slice_sample(n = 50000, replace = TRUE)

ggplot(meth_sampled, aes(x = Methylation, color = dx)) +
    geom_density(linewidth = 1) +
    labs(
        title = paste0("Methylation Distribution by Diagnosis: ", region_name),
        x = "Methylation Proportion (Beta)",
        y = "Density",
        color = "Diagnosis"
    ) +
    scale_color_brewer(palette = "Set2")
```

### 5.4 Sample Correlation

Pairwise correlations between samples help identify outliers. Biological replicates and similar sample types should have high correlations (typically >0.9).

```{r}
#| label: correlation-matrix

## Compute correlation matrix (on subset for efficiency)
cor_matrix <- cor(meth_matrix, use = "pairwise.complete.obs")

## Summary statistics
cor_values <- cor_matrix[upper.tri(cor_matrix)]
cat("Correlation Summary:\n")
cat("  Minimum:", round(min(cor_values), 3), "\n")
cat("  Median:", round(median(cor_values), 3), "\n")
cat("  Maximum:", round(max(cor_values), 3), "\n")
```

```{r}
#| label: fig-correlation-heatmap
#| fig-cap: "Sample correlation heatmap. High correlations (yellow) indicate similar methylation profiles. Samples with consistently low correlations may be outliers."
#| fig-width: 10
#| fig-height: 9

## Create annotation for heatmap
annotation_df <- phenotype_data %>%
    filter(brnum %in% colnames(cor_matrix)) %>%
    select(brnum, dx, sex) %>%
    column_to_rownames("brnum")

## Plot heatmap
pheatmap(cor_matrix,
         color = colorRampPalette(brewer.pal(9, "YlOrRd"))(100),
         annotation_col = annotation_df,
         annotation_row = annotation_df,
         show_rownames = FALSE,
         show_colnames = FALSE,
         main = paste0("Sample Correlations: ", region_name))
```

#### Identify Potential Outliers

```{r}
#| label: outlier-detection

## Calculate mean correlation for each sample
mean_cor <- rowMeans(cor_matrix) - 1/(ncol(cor_matrix))  # Exclude self-correlation

## Flag samples with unusually low correlation
cor_threshold <- mean(mean_cor) - 2 * sd(mean_cor)

outlier_samples <- tibble(
    Sample = names(mean_cor),
    Mean_Correlation = mean_cor
) %>%
    filter(Mean_Correlation < cor_threshold)

if (nrow(outlier_samples) > 0) {
    cat("Potential outliers (mean correlation > 2 SD below mean):\n")
    print(outlier_samples)
} else {
    cat("No correlation-based outliers detected.\n")
}
```

### 5.5 Principal Component Analysis

PCA reduces the high-dimensional methylation data to a few principal components, helping visualize sample relationships and identify batch effects.

```{r}
#| label: pca-analysis

## Remove rows with missing values for PCA
meth_complete <- meth_matrix[complete.cases(meth_matrix), ]

## Run PCA
pca_result <- prcomp(t(meth_complete), scale. = TRUE, center = TRUE)

## Calculate variance explained
var_explained <- summary(pca_result)$importance[2, ] * 100

cat("Variance explained by PCs:\n")
cat("  PC1:", round(var_explained[1], 1), "%\n")
cat("  PC2:", round(var_explained[2], 1), "%\n")
cat("  PC3:", round(var_explained[3], 1), "%\n")
```

```{r}
#| label: fig-pca-diagnosis
#| fig-cap: "PCA colored by diagnosis. Separation along PC1/PC2 may indicate biological differences or technical artifacts."
#| fig-width: 8
#| fig-height: 6

## Prepare PCA data for plotting
pca_df <- as.data.frame(pca_result$x) %>%
    rownames_to_column("Sample") %>%
    left_join(phenotype_data %>% select(brnum, dx, sex, agedeath),
              by = c("Sample" = "brnum"))

ggplot(pca_df, aes(x = PC1, y = PC2, color = dx)) +
    geom_point(size = 3, alpha = 0.8) +
    labs(
        title = paste0("PCA by Diagnosis: ", region_name),
        x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
        y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
        color = "Diagnosis"
    ) +
    scale_color_brewer(palette = "Set2") +
    theme(legend.position = "right")
```

```{r}
#| label: fig-pca-sex
#| fig-cap: "PCA colored by sex. Strong separation by sex may indicate chromosome X effects or batch effects."
#| fig-width: 8
#| fig-height: 6

ggplot(pca_df, aes(x = PC1, y = PC2, color = sex)) +
    geom_point(size = 3, alpha = 0.8) +
    labs(
        title = paste0("PCA by Sex: ", region_name),
        x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
        y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
        color = "Sex"
    ) +
    scale_color_brewer(palette = "Set1") +
    theme(legend.position = "right")
```

```{r}
#| label: fig-pca-age
#| fig-cap: "PCA colored by age. Gradients may indicate age-related methylation changes."
#| fig-width: 8
#| fig-height: 6

ggplot(pca_df, aes(x = PC1, y = PC2, color = agedeath)) +
    geom_point(size = 3, alpha = 0.8) +
    labs(
        title = paste0("PCA by Age: ", region_name),
        x = paste0("PC1 (", round(var_explained[1], 1), "%)"),
        y = paste0("PC2 (", round(var_explained[2], 1), "%)"),
        color = "Age at Death"
    ) +
    scale_color_viridis_c(option = "plasma") +
    theme(legend.position = "right")
```

---

## 6. CpG Filtering

Based on the QC results, we apply coverage-based filtering to retain only high-quality CpG sites.

```{r}
#| label: cpg-filtering

## Define filtering criteria
min_coverage <- 5        # Minimum mean coverage per CpG
min_sample_frac <- 0.8   # CpG must be covered in at least 80% of samples

## Calculate coverage per CpG across samples
cov_per_cpg <- getCoverage(bsseq_data)
samples_with_coverage <- rowSums(cov_per_cpg >= min_coverage, na.rm = TRUE)
fraction_covered <- samples_with_coverage / ncol(bsseq_data)

## Apply filter
keep_cpgs <- fraction_covered >= min_sample_frac

cat("CpG Filtering Summary:\n")
cat("  Original CpGs:", format(nrow(bsseq_data), big.mark = ","), "\n")
cat("  CpGs passing filter:", format(sum(keep_cpgs), big.mark = ","), "\n")
cat("  CpGs removed:", format(sum(!keep_cpgs), big.mark = ","), "\n")
cat("  Retention rate:", round(mean(keep_cpgs) * 100, 1), "%\n")
```

---

## 7. QC Summary Table

```{r}
#| label: qc-summary-table

## Compile QC metrics for each sample
qc_summary <- tibble(
    Sample = colnames(bsseq_data),
    Mean_Coverage = mean_coverage[colnames(bsseq_data)],
    Mean_Correlation = mean_cor[colnames(bsseq_data)],
    Global_Methylation = global_meth[colnames(bsseq_data)]
) %>%
    left_join(phenotype_data %>% select(brnum, dx, sex, agedeath),
              by = c("Sample" = "brnum")) %>%
    mutate(
        Coverage_Flag = ifelse(Mean_Coverage < 5, "Low", "OK"),
        Correlation_Flag = ifelse(Mean_Correlation < cor_threshold, "Outlier", "OK"),
        QC_Status = case_when(
            Coverage_Flag == "Low" | Correlation_Flag == "Outlier" ~ "Review",
            TRUE ~ "Pass"
        )
    )

## Display summary
knitr::kable(
    qc_summary %>%
        select(Sample, dx, Mean_Coverage, Mean_Correlation, QC_Status) %>%
        mutate(
            Mean_Coverage = round(Mean_Coverage, 1),
            Mean_Correlation = round(Mean_Correlation, 3)
        ),
    caption = "QC Summary Table"
)
```

### Samples Requiring Review

```{r}
#| label: samples-review

review_samples <- qc_summary %>% filter(QC_Status == "Review")

if (nrow(review_samples) > 0) {
    cat("The following samples may require additional review:\n\n")
    print(review_samples %>% select(Sample, dx, Mean_Coverage, Mean_Correlation, Coverage_Flag, Correlation_Flag))
} else {
    cat("All samples passed QC. No samples require additional review.\n")
}
```

---

## 8. Save QC Results

```{r}
#| label: save-results

## Create output directory
output_dir <- here("qc_analysis/_m")
if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
}

## Save QC summary table
write.csv(qc_summary,
          file.path(output_dir, paste0("qc_summary_", region, ".csv")),
          row.names = FALSE)

## Save filtered BSseq object
bsseq_filtered <- bsseq_data[keep_cpgs, ]
out_h5se <- file.path(output_dir, paste0(region, "_bsseq_qc_filtered_h5se"))
if (dir.exists(out_h5se)) unlink(out_h5se, recursive = TRUE)
saveHDF5SummarizedExperiment(bsseq_filtered, dir = out_h5se, replace = TRUE)

cat("Results saved to:", output_dir, "\n")
cat("  - QC summary: qc_summary_", region, ".csv\n", sep = "")
cat("  - Filtered BSseq: ", region, "_bsseq_qc_filtered_h5se/\n", sep = "")
```

---

## 9. Exercises

Test your understanding with these questions:

1. **Coverage interpretation**: Which samples have the lowest coverage? Would you recommend removing any samples based on coverage alone?

2. **Methylation distributions**: Do the methylation distributions look bimodal as expected? Are there any samples with unusual patterns?

3. **Batch effects**: Based on the PCA plots, do you see evidence of batch effects (e.g., clustering by processing date rather than biology)?

4. **Threshold sensitivity**: Re-run the CpG filtering with a 10x coverage threshold instead of 5x. How does this affect the number of retained CpGs?

5. **Outlier investigation**: If you identified any outlier samples, what additional information would you want to check before deciding to remove them?

6. **Expand to genetic similarity**: Expand QC correlation from age and sex to genetic similarity (snpPCs 1-3) within phenotype data.

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
