---
title: "WGBS Quality Control: Hippocampus"
author: "HEART-GeN Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
    self-contained: true
execute:
  warning: false
  message: false
---

## Learning Objectives

By the end of this analysis, you will be able to:

- Load and explore pre-generated WGBS data for the hippocampus
- Calculate and interpret key QC metrics for methylation data
- Identify samples and CpG sites that may need filtering
- Create publication-quality QC figures
- Make informed decisions about data quality thresholds

**Note:** For detailed background on WGBS methodology and QC concepts, see the companion tutorial `wgbs_qc_bsseq.qmd`.

---

## 1. Background

The **hippocampus** is a brain structure critical for learning, memory formation, and spatial navigation. Hippocampal abnormalities are commonly observed in schizophrenia, including reduced volume and altered gene expression, making it an important region for epigenetic studies.

In this analysis, we will perform quality control on WGBS data from hippocampal samples, focusing on:

- **Coverage**: Are CpG sites adequately sequenced?
- **Methylation distributions**: Do patterns look biologically plausible?
- **Sample correlations**: Are replicates similar to each other?
- **Principal components**: Are there technical batch effects?

---

## 2. Environment Setup

### Activate Conda Environment

Before running this analysis, ensure you have activated the lab's epigenomics environment:
```bash
source /projects/p32505/opt/miniforge3/etc/profile.d/conda.sh
conda activate /projects/p32505/opt/envs/epigenomics
```

### Load Libraries

```{r}
#| label: setup
#| message: false

library(here)
library(bsseq)
library(HDF5Array)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(pheatmap)
library(RColorBrewer)

## Set ggplot theme for consistency
theme_set(theme_bw(base_size = 12))

## Define region for this analysis
region <- "hippocampus"
region_name <- "Hippocampus"
```

---

## 3. Data Loading

### Load Pre-Generated BSseq Object

The combined BSseq object was generated using the data loading scripts. This object contains methylation data for all chromosomes combined into a single, HDF5-backed structure for efficient access.

```{r}
#| label: load-data

## Path to the combined data
data_path <- here(/path/to/data/) # Explore "inputs/"

## Load the HDF5-backed BSseq object
bsseq_data <- loadHDF5SummarizedExperiment(
    file.path(data_path, paste0(region, "_bsseq_h5se"))
)

## Load phenotype data
phenotype_data <- read.csv(
    file.path(data_path, paste0(region, "_phenotypes.csv")),
    stringsAsFactors = FALSE
)

## Display basic information
cat("BSseq object summary:\n")
cat("  Number of CpG sites:", nrow(bsseq_data), "\n")
cat("  Number of samples:", ncol(bsseq_data), "\n")
cat("  Genome:", unique(genome(bsseq_data)), "\n")
```

### Phenotype Overview

We want to examine adults from our region of interest.

```{r}
#| label: phenotype-overview

## Display phenotype summary
```

---

## 4. Sample Overview

### Sample Metadata Table

```{r}
#| label: sample-table

## Create sample summary table

## Display first few samples
```

---

## 5. QC Analysis

### 5.1 Sequencing Coverage

Coverage refers to the number of sequencing reads that overlap each CpG site. Higher coverage provides more reliable methylation estimates.

#### Calculate Per-Sample Coverage

```{r}
#| label: coverage-calculation

## Get coverage matrix (reads per CpG per sample)

## Calculate mean coverage per sample

## Create coverage summary dataframe

## Summary statistics (e.g., minimum, median, maxium)
cat("Coverage Summary:\n")
```

#### Coverage Barplot

```{r}
#| label: fig-coverage-barplot
#| fig-cap: "Mean sequencing coverage per sample. Samples below the red dashed line (5x coverage) may have unreliable methylation estimates."
#| fig-width: 10
#| fig-height: 6

```

#### Identify Low-Coverage Samples

```{r}
#| label: low-coverage-samples

## Flag samples with coverage < 5x
```

### 5.2 CpG Site Coverage

Not all CpG sites are equally well-covered. Sites with very low coverage across samples should be filtered before analysis.

```{r}
#| label: cpg-coverage

## Calculate coverage across CpGs

## Calculate percentage of CpGs meeting thresholds

## Output table using knitr::kable
```

```{r}
#| label: fig-cpg-coverage-hist
#| fig-cap: "Distribution of mean coverage across CpG sites. Most sites should have adequate coverage (>5x)."
#| fig-width: 8
#| fig-height: 5

## Plot distribution (capped at 50x for visualization)
```

### 5.3 Methylation Distribution

WGBS data typically shows a **bimodal distribution** with peaks near 0% (unmethylated) and 100% (fully methylated) methylation.

```{r}
#| label: methylation-extraction

## Extract methylation proportions (beta values)
## Use a subset of CpGs for efficiency (should be reproducible)

```

#### Per-Sample Methylation Histograms

```{r}
#| label: fig-meth-histograms
#| fig-cap: "Methylation distribution for selected samples. A bimodal pattern (peaks at 0 and 1) is expected for high-quality WGBS data."
#| fig-width: 10
#| fig-height: 8

## Select a few representative samples

## Prepare data for plotting

## Plot the data

```

#### Density Plot by Diagnosis

```{r}
#| label: fig-meth-density
#| fig-cap: "Methylation density by diagnosis group. Distributions should be similar across groups if there are no technical artifacts."
#| fig-width: 8
#| fig-height: 5

## Calculate global methylation per sample

## Create density data with diagnosis

## Sample for plotting efficiency

```

### 5.4 Sample Correlation

Pairwise correlations between samples help identify outliers. Biological replicates and similar sample types should have high correlations (typically >0.9).

```{r}
#| label: correlation-matrix

## Compute correlation matrix (on subset for efficiency)

## Summary statistics
cat("Correlation Summary:\n")

```

```{r}
#| label: fig-correlation-heatmap
#| fig-cap: "Sample correlation heatmap. High correlations (yellow) indicate similar methylation profiles. Samples with consistently low correlations may be outliers."
#| fig-width: 10
#| fig-height: 9

## Create annotation for heatmap

## Plot heatmap

```

#### Identify Potential Outliers

```{r}
#| label: outlier-detection

## Calculate mean correlation for each sample

## Flag samples with unusually low correlation
```

### 5.5 Principal Component Analysis

PCA reduces the high-dimensional methylation data to a few principal components, helping visualize sample relationships and identify batch effects.

```{r}
#| label: pca-analysis

## Remove rows with missing values for PCA

## Run PCA

## Calculate variance explained

## Print the top 3-5 PCs
cat("Variance explained by PCs:\n")

```

```{r}
#| label: fig-pca-diagnosis
#| fig-cap: "PCA colored by diagnosis. Separation along PC1/PC2 may indicate biological differences or technical artifacts."
#| fig-width: 8
#| fig-height: 6

## Prepare PCA data for plotting

## Plot data
```

```{r}
#| label: fig-pca-sex
#| fig-cap: "PCA colored by sex. Strong separation by sex may indicate chromosome X effects or batch effects."
#| fig-width: 8
#| fig-height: 6

## Examine PCA colored by sex
```

```{r}
#| label: fig-pca-age
#| fig-cap: "PCA colored by age. Gradients may indicate age-related methylation changes."
#| fig-width: 8
#| fig-height: 6

## Examine PCA by age
```

---

## 6. CpG Filtering

Based on the QC results, we apply coverage-based filtering to retain only high-quality CpG sites.

```{r}
#| label: cpg-filtering

## Define filtering criteria
min_coverage <- 5        # Minimum mean coverage per CpG
min_sample_frac <- 0.8   # CpG must be covered in at least 80% of samples

## Calculate coverage per CpG across samples

## Apply filter

## Summarize results (before and after filtering; how many CpGs were retained, rate of retention)
cat("CpG Filtering Summary:\n")

```

---

## 7. QC Summary Table

```{r}
#| label: qc-summary-table

## Compile QC metrics for each sample

## Display summary
```

### Samples Requiring Review

```{r}
#| label: samples-review

## What are the samples that require review
```

---

## 8. Save QC Results

```{r}
#| label: save-results

## Create output directory
output_dir <- here("qc_analysis/_m")
if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
}

## Save QC summary table

## Save filtered BSseq object

```

---

## 9. Exercises

Test your understanding with these questions:

1. **Coverage interpretation**: Which samples have the lowest coverage? Would you recommend removing any samples based on coverage alone?

2. **Methylation distributions**: Do the methylation distributions look bimodal as expected? Are there any samples with unusual patterns?

3. **Batch effects**: Based on the PCA plots, do you see evidence of batch effects (e.g., clustering by processing date rather than biology)?

4. **Threshold sensitivity**: Re-run the CpG filtering with a 10x coverage threshold instead of 5x. How does this affect the number of retained CpGs?

5. **Outlier investigation**: If you identified any outlier samples, what additional information would you want to check before deciding to remove them?

6. **Expand to genetic similarity**: Expand QC correlation from age and sex to genetic similarity (snpPCs 1-3) within phenotype data.

---

## Session Info

```{r}
#| label: session-info

sessionInfo()
```
