---
title: "Quality Control of Whole-Genome Bisulfite Sequencing (WGBS) Data"
author: "HEART-GeN Lab"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: cosmo
execute:
  warning: false
  message: false
---

## Learning Objectives

By the end of this tutorial, you will be able to:

- Explain what WGBS data measures
- Load methylation data into R using `bsseq`
- Perform standard QC checks for WGBS data
- Identify low-quality samples and CpG sites
- Interpret plots commonly used in methylation QC

---

## Background: What is WGBS?

Whole-Genome Bisulfite Sequencing (WGBS) measures **DNA methylation** at cytosine–phosphate–guanine (CpG) sites across the genome.

Each CpG site has:
- Number of **methylated reads**
- Number of **unmethylated reads**
- A **methylation proportion**:

\[
\text{Methylation} = \frac{\text{methylated reads}}{\text{total reads}}
\]

Because WGBS data is noisy and expensive, **quality control (QC)** is essential before downstream analysis.

### What is DNA Methylation?

#### Definition
DNA methylation is an **epigenetic modification** involving the addition of a **methyl group (-CH₃)** to DNA, typically at the **5th carbon of cytosine residues** within **CpG dinucleotides**.

- **CpG = Cytosine followed by Guanine** (linked by a phosphate bond on the DNA backbone).  
- After methylation, the base becomes **5-methylcytosine (5mC)**.

#### Biological Roles
- **Gene regulation:** Methylation in gene promoters often *silences* gene expression by preventing transcription factor binding.  
- **Genomic stability:** Suppresses repetitive elements and transposable elements.  
- **Cell identity:** Helps cells maintain their specialized functions during development.  
- **Imprinting and X-inactivation:** Controls allele-specific expression and dosage compensation.

#### Key Idea
Think of DNA methylation as a *sticky note* added to DNA — it doesn’t change the code, but gives cells extra instructions about which genes to read or ignore.

### Measuring DNA Methylation

Several technologies can detect DNA methylation, such as:
- **WGBS (Whole-Genome Bisulfite Sequencing)**
- **RRBS (Reduced Representation Bisulfite Sequencing)**
- **Methylation arrays** (e.g., Illumina 450k, EPIC)
- **Enzyme-based methods** (e.g., MeDIP-seq, MRE-seq)

Among these, **WGBS** is the **gold standard** for base-resolution, genome-wide measurement of methylation.

### Principle of Bisulfite Conversion

#### Core Reaction
- **Sodium bisulfite** treatment chemically converts **unmethylated cytosines** into **uracil**, which is then read as thymine after PCR and sequencing.  
- **Methylated cytosines** (5mC) are *protected* and remain as cytosines.

| Base in Original DNA | After Bisulfite | Sequencing Read |
|----------------------|-----------------|-----------------|
| Unmethylated C       | U → T            | T               |
| Methylated C (5mC)   | (no change)      | C               |

This simple but powerful chemistry allows you to **infer methylation status** at single-nucleotide resolution.

### What is Whole-Genome Bisulfite Sequencing (WGBS)?

#### Overview
WGBS is a sequencing-based method that measures cytosine methylation **across the entire genome** using **bisulfite conversion followed by next-generation sequencing**.

#### General Workflow
1. **Extract DNA**
2. **Fragment DNA** into manageable pieces (~200–300 bp)
3. **Bisulfite convert** to distinguish methylated from unmethylated cytosines
4. **Library preparation** (adapter ligation, PCR amplification)
5. **Sequencing** (using Illumina or similar platforms)
6. **Alignment to reference genome** using bisulfite-aware algorithms (e.g., *Bismark*, *BS-Seeker2*)
7. **Methylation calling** (counting C vs T at CpG sites)
8. **Downstream analysis** (e.g., differential methylation, annotation, visualization)

### Data Interpretation

After sequencing and mapping:
- For each CpG site, WGBS provides a **methylation fraction**  
  \[
  \text{Methylation level} = \frac{\text{# of reads with C}}{\text{# of reads with C or T}}
  \]
- Values range from:
  - **0.0 (unmethylated)** → mostly T reads  
  - **1.0 (fully methylated)** → mostly C reads  

#### Example Output
| CpG site | C reads | T reads | Methylation (%) |
|-----------|----------|----------|-----------------|
| chr1:1050 | 18 | 2 | 90% |
| chr1:2050 | 4 | 16 | 20% |
| chr1:3050 | 0 | 20 | 0% |

### Advantages and Limitations

#### Advantages
- **Single-base resolution** across the genome.  
- **Unbiased** (covers both CpG-rich and CpG-poor regions).  
- Enables integrative analysis with RNA-seq, ATAC-seq, etc.

#### Limitations
- **Expensive** due to deep coverage requirements.  
- **DNA degradation** during bisulfite treatment.  
- **Computational complexity:** mapping and QC are nontrivial.  
- **Does not distinguish 5mC from 5hmC** (hydroxymethylcytosine) without secondary methods.

### Applications in Biology and Medicine

- **Developmental biology:** tracking methylation during cell differentiation.
- **Cancer epigenomics:** identifying tumor-specific methylation changes.
- **Neurodegeneration:** studying altered epigenetic patterns in Alzheimer’s or Parkinson’s disease.
- **Evolutionary biology:** comparing epigenetic landscapes across species.

### Key Computational Tools

| Task | Common Tools |
|------|---------------|
| Read trimming | Trim Galore!, cutadapt |
| Alignment | Bismark, BS-Seeker2, bwa-meth |
| Methylation calling | MethylDackel, Bismark methylation extractor |
| DMR analysis | DSS, methylKit, bsseq |
| Visualization | IGV (bisulfite mode), R packages (ggplot2, genomation) |

### Summary Points

- DNA methylation is a key **epigenetic regulator** of gene expression.  
- **Bisulfite conversion** provides the chemical basis for distinguishing methylated and unmethylated cytosines.  
- **WGBS** offers nucleotide-level methylation mapping of the whole genome, though it is resource-intensive.  
- Data interpretation often requires integrative bioinformatics workflows and careful quality control.

---

## Software Requirements and Conda Environments

This tutorial uses **R** and **Bioconductor**.

### Conda Environments
For this work, Dr. Benjamin has already created shared **Conda environments** on Quest that contain all the
packages needed for your analyses. You do **not** need to build your own environment; you only need to
configure your HPC session to use the lab environment and then activate it.

#### Background: Conda in this lab

- **Conda environments** are isolated software setups that bundle specific versions of Python/R and all required packages, ensuring analyses are reproducible and stable across users.  
- In this lab, shared environments are installed in a **project path on Quest** (not in your `$HOME`), so everyone uses the **same, curated** toolchain.  

#### Lab Conda environments on Quest

Below are the main shared environments available on Northwestern Quest for different analysis tasks:

- `rnaseq`: Bulk RNA-seq analysis (edgeR, limma, etc.).  
- `epigenomics`: DNA methylation and other epigenomic analyses (e.g. BSSeq and related tools).  
- `ml`: GPU-based machine learning (CuPy, RAPIDS, PyTorch, scikit-learn, JAX).  
- `network`: Network and systems biology (WGCNA, igraph, networkx, GNVAE/BSVAE deps).  
- `genomics`: GWAS/xQTL, variant calling, and general sequence tools (tensorqtl, localqtl, RFMix-reader, etc.).  
- `software-dev`: General development and testing; **not for final analyses**.  

For this project, you will specifically use the **`epigenomics`** environment.

#### Set up your PATH in `.bashrc`

Add the following line to your `~/.bashrc` (or `~/.bash_profile` if that is what your shell uses):

```{bash}
export PATH=~/.local/bin:/projects/p32505/opt/bin:/projects/p32505/opt/miniforge3/bin:$PATH
```

Then reload your shell configuration:

```{bash}
source ~/.bashrc
```

This ensures:
- User-level scripts in `~/.local/bin` are found first.  
- Lab-wide tools in `/projects/p32505/opt/bin` are available.  
- The lab’s Miniforge/Conda installation in `/projects/p32505/opt/miniforge3/bin` is on your PATH so `conda` works without installing anything in `$HOME`.  

#### Using the epigenomics Conda environment

After updating your PATH and starting a new shell (or running `source ~/.bashrc`):

```{bash}
conda activate /projects/p32505/opt/envs/epigenomics
```

Guidelines:
- Use this environment **as-is** for all DNA methylation/epigenomics work.  
- Do **not** run `conda install` or `pip install` in this shared environment; request changes through the lab’s usual process.

#### Important notes

- Do **not** use `conda install`, `pip install`, or other package managers inside `/projects/p32505/opt/envs/...`.  
- Use the environment **as-is** for analysis; environment updates are handled centrally (via `environment.yml`, lockfiles, and a rebuild process) to keep results reproducible.  
- Always double-check that `epigenomics` is active before running pipelines or R/Python scripts for methylation work.

### Load libraries

```{r}
library(bsseq)
library(ggplot2)
```

---

## Loading Example WGBS Data

We will use an example dataset included with `bsseq`.

```{r}
data(BS.cancer.ex)
BS.cancer.ex
```

**Interpretation:**

* Rows = CpG sites
* Columns = samples
* Each CpG has methylated and total read counts

---

## QC Step 1: Sequencing Coverage

### Why coverage matters

Coverage matters because methylation is estimated from a limited number of reads, and low coverage makes those estimates noisy, unstable, and misleading.

#### What does "coverage" mean?

At a single CpG site, **coverage** is the total number of sequencing reads that overlap that position.

Example:

| CpG site | Methylated reads | Unmethylated reads | Coverage |
| -------- | ---------------- | ------------------ | -------- |
| CpG_1    | 3                | 2                  | 5×       |
| CpG_2    | 18               | 2                  | 20×      |

Coverage = methylated + unmethylated reads.

#### How methylation is estimated

For each CpG, methylation is calculated as:

[
\text{Methylation proportion} =
\frac{\text{methylated reads}}{\text{total reads}}
]

This is an **estimate**, not a true value. Like all estimates, it becomes **more reliable as the sample size increases**.

#### Low coverage = high uncertainty

Think of methylation as a **coin-flipping experiment**:

* "Heads" = methylated read
* "Tails" = unmethylated read

If you flip a coin:

* **5 times**, the result might be very misleading
* **50 times**, the estimate stabilizes

**Example:**

| Coverage | Methylated reads | Estimated methylation |
| -------- | ---------------- | --------------------- |
| 2×       | 2                | 100%                  |
| 4×       | 3                | 75%                   |
| 20×      | 15               | 75%                   |

At **2× coverage**, 100% methylation looks extreme — but it’s based on only **two reads**.

At **20× coverage**, 75% methylation is much more trustworthy.

#### Statistical intuition

* Each read is a **random sample**
* Low coverage → high randomness
* High coverage → averaging out noise

This means:

* Low-coverage CpGs **jump around**
* High-coverage CpGs are **stable**

#### Low coverage exaggerates differences between samples

Suppose two samples are biologically identical.

At low coverage:

| Sample | Reads | Methylated | Estimated methylation |
| ------ | ----- | ---------- | --------------------- |
| A      | 3×    | 3          | 100%                  |
| B      | 3×    | 1          | 33%                   |

They look **dramatically different**, but the difference is just **sampling noise**.

At higher coverage:

| Sample | Reads | Methylated | Estimated methylation |
| ------ | ----- | ---------- | --------------------- |
| A      | 20×   | 14         | 70%                   |
| B      | 22×   | 15         | 68%                   |

Now the estimates agree.

**Low coverage creates false biological differences.**

#### Low coverage inflates false positives

In downstream analysis (e.g. differential methylation):

* Low-coverage CpGs appear highly variable
* High variability increases false positives
* Statistical tests become unreliable

This is why most pipelines:

* **Filter CpGs by coverage**
* Or **downweight low-coverage sites**

#### Coverage affects smoothing and regional analysis

WGBS data is often **smoothed across nearby CpGs**.

Low-coverage CpGs:

* Contribute noisy values
* Distort regional averages
* Reduce power to detect true DMRs

High-coverage CpGs:

* Anchor the signal
* Improve regional consistency

#### Sample-level coverage matters too

Coverage is not just a CpG issue — it’s also a **sample QC metric**.

If one sample has:

* Much lower average coverage
* More missing CpGs
* Higher noise

Then:

* PCA will separate it
* Correlations will drop
* It may dominate technical variation

Such samples are often **removed entirely**.

#### Why common cutoffs are 3×, 5×, or 10×

There is no single "correct" cutoff, but common rules of thumb:

| Coverage cutoff | Use case                       |
| --------------- | ------------------------------ |
| ≥3×             | Very exploratory analysis      |
| ≥5×             | Standard QC filtering          |
| ≥10×            | High-confidence analysis       |
| ≥20×            | Very strict / targeted studies |

Trade-off:

* Higher cutoff → cleaner data
* Higher cutoff → fewer CpGs retained

### Calculate average coverage per sample

```{r}
coverage <- colMeans(getCoverage(BS.cancer.ex))
coverage
```

### Plot average coverage

```{r}
barplot(
    coverage, main = "Average Coverage per Sample",
    ylab = "Reads per CpG", las = 2
)
```

**Guidelines:**

* `< 5×` coverage → poor quality
* `10–30×` coverage → typical
* Large differences between samples may indicate technical issues

---

## QC Step 2: Methylation Distribution

### Why check methylation distributions?

High-quality WGBS data usually shows:

* Many CpGs near **0% methylation**
* Many CpGs near **100% methylation**
* Fewer CpGs in between

### Extract methylation proportions

```{r}
meth <- getMeth(BS.cancer.ex, type = "raw")
```

### Plot histogram for one sample

```{r}
hist(
    meth[,1], breaks = 50, col = "steelblue",
    main = "Methylation Distribution (Sample 1)",
    xlab = "Methylation proportion"
)
```
**Interpretation:**

* A **bimodal** distribution is expected
* Flat or noisy distributions may indicate technical problems

---

## QC Step 3: Sample-to-Sample Correlation

### Why correlations matter

Biological replicates should be more similar to each other than unrelated samples.

### Compute correlation matrix

```{r}
cor_matrix <- cor(meth, use = "pairwise.complete.obs")
cor_matrix
```

### Visualize correlations

```{r}
heatmap(cor_matrix, symm = TRUE, main = "Sample Correlation Heatmap")
```

**Interpretation:**

* Correlations > 0.9 are typical
* Samples with consistently low correlations are potential outliers

---

## QC Step 4: Principal Component Analysis (PCA)

### Why PCA?

PCA helps identify:

* Outlier samples
* Batch effects
* Unexpected clustering

### Run PCA

```{r}
pca <- prcomp(t(meth), scale. = TRUE)
```

### Plot PCA results

```{r}
plot(
    pca$x[,1], pca$x[,2], xlab = "PC1", ylab = "PC2",
    pch = 19, col = "darkred"
)

text(pca$x[,1], pca$x[,2], labels = colnames(meth), pos = 3)
```

**Interpretation:**

* Samples clustering together are similar
* Isolated samples may have QC issues

---

## Filtering Low-Quality CpG Sites

### Remove low-coverage CpGs

```{r}
BS.filtered <- BS.cancer.ex[rowMeans(getCoverage(BS.cancer.ex)) >= 5,]
BS.filtered
```

### Why filtering is important

Filtering:

* Reduces noise
* Improves statistical power
* Makes downstream analyses more reliable

---

## Summary of WGBS QC Workflow

**QC checklist:**

1. Check sequencing depth
2. Inspect methylation distributions
3. Examine sample correlations
4. Identify outliers with PCA
5. Filter low-coverage CpG sites

---

## Student Exercises

1. Identify the sample with the lowest average coverage.
2. Redo the PCA after removing one sample.
3. Compare methylation distributions before and after CpG filtering.
4. Try a different coverage cutoff (e.g., 3× or 10×).

---

## Key Takeaways

* QC is essential before any WGBS analysis
* Poor-quality samples can distort biological conclusions
* Visualization is one of the most powerful QC tools in genomics

---

## References

* `bsseq` Bioconductor vignette: https://www.bioconductor.org/packages/devel/bioc/vignettes/bsseq/inst/doc/bsseq.html
* Hansen et al., *Genome Biology* (2012)
